taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-08-12 paper/manuscript.Rmd
Change some order of ideas, remove old comments.

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
#  - name: "Jorrit Poelen"
#    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a consistent, standardized layout or schema (discussed below). These files are automatically downloaded and imported and stored on a local database by `taxadb`.  The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes under a GB, and can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.  In contrast, querying a single name over the web API, requiring the server to respond, execute the query, and serialize the response, can take several seconds. Thus it does not take many taxa before transferring the entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup and enabling offline use.

@@

It is not necessary to call `td_create()`.  Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.

@@

KN: I don't know if this section needs to be in here, especially since `synonym()` doesn't fit cleanly in the four families of functions described at the beginning (maybe it's a helper function?). Can instead use the unresolved names discussion to segue into `by_` functions like in the tutorial. Something along the lines of "A name unresolved with `get_ids` could be due to misspelling, a missing synonym in the provider database OR a name resolving to multiple synonyms - 'by_' gives you access to all the synonym options." When it's laid out like that it also makes 'synonym()' seem less necessary.

@@

@@

```{r}
most_synonyms$n
```
Such queries are also not possible through the API or web interface to Catalogue of Life. To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.  This should be done with care as an unfiltered table may require several gigabytes of memory to load.

Some synonyms may also be associated with more than one unique identifier.  This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.  The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.  Other functions in `taxadb` allow us to explore these names further.

Having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:

>>>>

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
#  - name: "Jorrit Poelen"
#    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
# - code: jorrit
#   address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files following a consistent, standardized layout or schema (discussed below). These files are automatically downloaded and imported and stored on a local database by `taxadb`.  The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes under a GB, and can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.  In contrast, querying a single name over the web API, requiring the server to respond, execute the query, and serialize the response, can take several seconds. Thus it does not take many taxa before transferring the entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup and enabling offline use.

@@

@@

@@

Some synonyms may also be associated with more than one unique identifier.  This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.  The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.

Other functions in `taxadb` allow us to explore these names further, for example having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:

@@

```{r}
most_synonyms$n
```

Such queries are also not possible through the API or web interface to Catalogue of Life. To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.  This should be done with care as an unfiltered table may require several gigabytes of memory to load.
